### 2023-02-23 doc/manuscript/translocation.qmd
Changed formatting option to wrap text by sentences
Add table of viability for early-late translocations

Reintroductions of *R. sierrae* were conducted in Yosemite National Park (California, USA) as part of an effort to reestablish populations of this endangered species. These reintroductions involved the collection of adult frogs from large, Bd-positive donor populations that have recovered in recent years following Bd-caused declines [@knapp2016] and translocating them to nearby recipient sites. We refer to these reintroductions as "translocations". During the period `r min(d1$year)`--`r max(d1$year)`, we conducted `r n_distinct(d1$site_id, d1$date)` translocations of frogs to `r n_distinct(d1$site_id)` recipient sites (`r transno_min`--`r transno_max` translocations per site; @fig-yosemap) <!--# How to restart numbering for figures in SI, add "S" prefix? -->. Each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs collected from one of `r n_distinct(d1$donor)` donor populations (@fig-yosemap). Estimates of frog survival during the one year following translocation indicate that survival was highly variable between recipient sites, but relatively constant within recipient sites (for the subset of sites that received multiple translocations; @fig-translocation-survival). As such, the average survival of frogs in the first translocation to a site was an important predictor of survival of frogs in all subsequent translocations to that site (@fig-transsurvival-postdens; see **Materials and Methods - Frog translocations** for analysis details), and suggests an important effect of site characteristics and/or (site \* frog) interactions on frog survival. For translocated frogs, Bd load changed little between immediately prior to translocation versus during the one year after translocation (@fig-bdload-beforeafter; see SI for analysis details). <!--# Add text describing that loads were below lethal threshold -->

To identify drivers of post-translocation frog survival, we developed a multilevel model in a Bayesian framework (see **Materials and Methods - Frog translocations** for analysis details). Our specific interest was whether Bd load had an important effect on adult survival, as would be expected if the effect of frog resistance was insufficient to overcome the negative effect of Bd. The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation was not among them (bd_load; @fig-survival-postdens). Important predictors included winter severity in the year following translocation (snow_t1), site elevation (elevation), and donor population (donor_72996; @fig-cond-effects). Males had somewhat higher survival than females, but this effect was only marginally important (sex_male; @fig-cond-effects). In addition to Bd load, effects of frog size (length), winter severity in the year of translocation (snow_t), day of year on which a translocation was conducted (day), and translocation order (order_first) were all unimportant (@fig-survival-postdens). In summary, our analyses show that Bd load is not a predictor of the survival of translocated frogs, consistent with frogs in recovering populations having evolved sufficient resistance and/or tolerance to minimize the adverse effects of Bd infection. In addition, the small changes in Bd load following translocation and relatively high survival of frogds in the majority of translocated cohorts suggests that resistant frogs from recovering populations can maintain the benefits of resistance (i.e., low-moderate Bd loads, minimal Bd-induced mortality despite ongoing infection) in non-natal habitats. Finally, in three locations where longer mark-recapture time series allowed us to assess the survival of adults recruited to the population that are the progeny of translocated individuals, naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults <!--# Show results. -->. Collectively, these results show that putative resistance in non-natal habitats is maintained across generations of frogs.

@@

<!--# Present results suggesting what donor-specific characteristics are responsible for important effect of donor population? -->

<!--# Mention the improved survival in recent compared with early translocations resulting from our better-informed site selection? Would put the viability results in better context because most of the low-viability populatons are those resulting from early translocations. Mark W: Yes, I think this is important.  I think we want to drive home the point that i) Acquired, heritable resistance in donor populations could promote the long-term viability of translocated populations and b) but survival is still highly dependent on environment independent of Bd. -->

@@

#### Conclusions

<!--# Placeholder for our conclusions section -->

@@

@@

We estimated survival of translocated frogs, recruitment of new frogs into the adult population, and population size using a Bayesian open-population Jolly-Seber mark-recapture model with known additions (i.e., translocated cohorts) to the population, as implemented by the mrmr package [@joseph2019] and using R Statistical Software [v4.4.4; @rsoftware2022]. The model tracks the states of *M* individuals that comprise a superpopulation made up of real and pseudo-individuals [see @joseph2018 for details]. The possible states of individuals include "not recruited", "alive", and "dead". The possible observations of individuals include "detected" and "not detected". We assume that individuals that are in the "not recruited" or "dead" states are never detected (i.e., there are no mistakes in the individual PIT tag records). For all models, we used mrmr defaults for priors, number of chains (4), and warmup and post-warmup iterations (2000 for each). We evaluated convergence of the Markov chain Monte Carlo (MCMC) algorithm using trace plots and Gelman-Rubin statistics (Rhat). Code used to construct CMR datasets and fit models for each translocated population is available from a GitHub repository ([https://github.com/SNARL1/cmr-analysis](https://github.com/SNARL1/cmr-analysis.)).

@@

To identify important predictors of frog survival following translocation, we developed multilevel models in a Bayesian framework. We performed all analyses using R Statistical Software [v4.4.4; @rsoftware2022] and the rstanarm package [@rstanarm2022]. We started with models that included all relevant population-level effects and, when data contained possible hierarchical structure, we included group-level effects in the full model. Within each analysis, we evaluated model fit using visualizations of leave-one-out ("LOO") probability integral transformations [@gelman2013, @gabry2019], and compared fits of models with and without group-level effects, and with differing group-level effects using LOO cross-validation [@vehtari2016]. Both approaches utilized the loo package.

@@

<!-- Mark W: We should specify whether the default priors were uninformative, slightly informative, etc. -->

@@

@@

![For frogs translocated to each recipient site, Bd loads for the period immediately prior to translocation versus during the 1-year period after translocation.](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="450"}

@@

-   Despite several mechanisms by which recovery could occur (dispersal, evolution, phenotypic plasticity), population recovery following severe disease-induced declines has rarely (?) been documented in wildlife populations (Rabbits and myxomatosis in Australia, rabbits and hemorrhagic disease in AU (Mutze et al. 2015); trout and whirling disease: Miller and Vincent (2008); frogs in Australia, Central America, Sierra Nevada; sea stars: Moritsch and Raimondi 2018, other examples? Restrict to species of conservation concern?).

>>>>

Reintroductions of *R. sierrae* were conducted in Yosemite National Park (California, USA) as part of an effort to reestablish populations of this endangered species.
These reintroductions involved the collection of adult frogs from large, Bd-positive donor populations that have recovered in recent years following Bd-caused declines [@knapp2016] and translocating them to nearby recipient sites.
We refer to these reintroductions as "translocations".
During the period `r min(d1$year)`--`r max(d1$year)`, we conducted `r n_distinct(d1$site_id, d1$date)` translocations of frogs to `r n_distinct(d1$site_id)` recipient sites (`r transno_min`--`r transno_max` translocations per site; @fig-yosemap) <!--# How to restart numbering for figures in SI, add "S" prefix? -->.
Each translocated cohort included `r transfrogs_min` to `r transfrogs_max` frogs collected from one of `r n_distinct(d1$donor)` donor populations (@fig-yosemap).
Estimates of frog survival during the one year following translocation indicate that survival was highly variable between recipient sites, but relatively constant within recipient sites (for the subset of sites that received multiple translocations; @fig-translocation-survival).
As such, the average survival of frogs in the first translocation to a site was an important predictor of survival of frogs in subsequent translocations to that site (@fig-transsurvival-postdens; see **Materials and Methods - Frog translocations** for analysis details).
This suggests an important effect of site characteristics and/or (site \* frog) interactions on frog survival.
In addition, 1-year survival was higher for frogs translocated later in the study period than earlier (\@tbl-viability-earlylate; Table 2) <!--# Why isn't cross-referencing working for this table? -->, which we suggest resulted primarily from our improved ability over time to choose recipient sites with higher habitat quality for *R. sierrae*.
Finally, Bd loads changed little between immediately prior to translocation versus during the one year after translocation (@fig-bdload-beforeafter; see SI for analysis details), and were nearly always well below the level indicative of severe chytridiomycosis and associated frog mortality (5.8 ITS copies on a log~10~ scale; @vredenburg2010; @joseph2018). <!--# Add median Bd load, proportion below 5.8 copies threshold? The text repeats in part similar results in Methods -- Translocaiton. -->

To identify site and frog-specific drivers of post-translocation frog survival, we developed a multilevel model in a Bayesian framework (see **Materials and Methods - Frog translocations** for analysis details).
Our specific interest was whether Bd load had an important effect on adult survival, as would be expected if the effect of frog resistance was insufficient to overcome the negative effect of Bd.
The best model of 1-year frog survival identified several important predictors, but Bd load at the time of translocation was not among them (bd_load; @fig-survival-postdens).
Important predictors included winter severity in the year following translocation (snow_t1), site elevation (elevation), and donor population (donor_72996; @fig-cond-effects).
Males had somewhat higher survival than females, but this effect was only marginally important (sex_male; @fig-cond-effects).
In addition to Bd load, effects of frog size (length), winter severity in the year of translocation (snow_t), day of year on which a translocation was conducted (day), and translocation order (order_first) were all unimportant (@fig-survival-postdens).
In summary, our analyses show that Bd load is not a predictor of the survival of translocated frogs, consistent with frogs in recovering populations having evolved sufficient resistance and/or tolerance to minimize the adverse effects of Bd infection.
In addition, the small changes in Bd load following translocation and relatively high survival of frogs in a substantial number of translocated cohorts suggests that resistant frogs from recovering populations can maintain the benefits of resistance in non-natal habitats (i.e., low-moderate Bd loads, minimal Bd-induced mortality).
Finally, in three locations where longer mark-recapture time series allowed us to assess the survival of adults recruited to the population (i.e., progeny of translocated individuals), naturally recruited adults had equivalent or higher survival probabilities than the originally translocated adults <!--# Show results. -->.
This result shows that frog resistance in non-natal habitats is maintained across generations of frogs.

@@

<!--# Drive home the point that i) Acquired, heritable resistance in donor populations could promote the long-term viability of translocated populations and b) but survival is still highly dependent on environment independent of Bd. -->

@@

#### Discussion

@@

<!--# Discuss effect of donor population on survival of translocated frogs. Differences in Bd resistance? -->

@@

We estimated survival of translocated frogs, recruitment of new frogs into the adult population, and population size using a Bayesian open-population Jolly-Seber mark-recapture model with known additions (i.e., translocated cohorts) to the population, as implemented by the mrmr package [@joseph2019] and using R Statistical Software [v4.4.4; @rsoftware2022].
The model tracks the states of *M* individuals that comprise a superpopulation made up of real and pseudo-individuals [see @joseph2018 for details].
The possible states of individuals include "not recruited", "alive", and "dead".
The possible observations of individuals include "detected" and "not detected".
We assume that individuals that are in the "not recruited" or "dead" states are never detected (i.e., there are no mistakes in the individual PIT tag records).
For all models, we used mrmr defaults for priors, number of chains (4), and warmup and post-warmup iterations (2000 for each).
We evaluated convergence of the Markov chain Monte Carlo (MCMC) algorithm using trace plots and Gelman-Rubin statistics (Rhat).
Code used to construct CMR datasets and fit models for each translocated population is available from the following GitHub repository: [https://github.com/SNARL1/cmr-analysis](https://github.com/SNARL1/cmr-analysis.).

@@

To identify important predictors of frog survival following translocation, we developed multilevel models in a Bayesian framework.
We performed all analyses using R Statistical Software [v4.4.4; @rsoftware2022] and the rstanarm package [@rstanarm2022].
We used default, weakly informative priors.
Starting models included all relevant population-level effects and, when data contained possible hierarchical structure, we included group-level effects in the full model.
Within each analysis, we evaluated model fit using visualizations of leave-one-out ("LOO") probability integral transformations [@gelman2013, @gabry2019], and compared fits of models with and without group-level effects, and with differing group-level effects using LOO cross-validation [@vehtari2016].
Both approaches utilized the loo package.

@@

@@

![For frogs translocated to each recipient site, Bd loads for the period immediately prior to translocation versus during the 1-year period after translocation. <!--# Remove gray background and gridlines -->](figures/bdload_beforeafter.png){#fig-bdload-beforeafter width="450"}

@@

```{r}
#| label: tbl-viability-earlylate
#| tbl-cap: "Proportion of populations translocated early versus late in the study period (< 2013 or >= 2013), respectively) for which the probability of extinction over 50 years was 1 or < 1 (viability = 0 or 1, respectively)."
#| echo: FALSE
#| message: FALSE

viability_table <- read_csv(here::here("data", "clean", "frog_translocation_final.csv")) %>% 
  distinct(site_id, year) %>% 
  group_by(site_id) %>% 
  filter(year == min(year)) %>% 
  mutate(site_id = as.character(site_id),
         viability = if_else(site_id %in% c(70414, 70370, 70134, 70641, 70619, 70505), 0, 1),
         period = if_else(year < 2013, "early", "late")) %>% 
  group_by(period) %>% 
  count(period, viability) %>% 
  mutate(n_period = sum(n), 
         prop = (n / n_period)) %>% 
  mutate(across(prop, round, 3)) %>% 
  arrange(viability, period)

  knitr::kable(viability_table, "pipe", align = "lcccr", caption = )

```

@@

-   Despite several mechanisms by which recovery could occur (dispersal, evolution, phenotypic plasticity), population recovery following severe disease-induced declines has rarely (?) been documented in wildlife populations (Rabbits and myxomatosis in Australia, rabbits and hemorrhagic disease in AU (Mutze et al. 2015); trout and whirling disease: Miller and Vincent (2008); frogs in Australia, Central America, Sierra Nevada; sea stars: Moritsch and Raimondi 2018, other examples?
    Restrict to species of conservation concern?)
    .
